Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-10 paper/paper.tex
paper review.

\begin{abstract}
We consider  the problem of reducing  the memory required  to run lazy
first-order functional  programs. To do this, we  analyze programs for
liveness of  heap-allocated data.  The  accompanying garbage collector
is modified  to use the result  of the analysis to  preserve only live
data---a subset  of reachable data---during  garbage collection.  This
results  in an  increase  in the  garbage  reclaimed and  consequently
reduces  the memory footprint  of programs.  While this  technique has
already been shown to  yield benefits for eager first-order languages,
the lack of a statically determinable execution order and the presence
of  closures   at  run-time  poses  additional   challenges  for  lazy
languages.   This  requires  changes  both  in  the  earlier  liveness
analysis and also in the garbage collection scheme required to reclaim
memory that is not live.

@@

\keywords{Heap Analysis, Memory Management, Garbage Collection}

@@

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The analysis  has two  parts.  First,  each  function is
summarized in a context-sensitive manner as a {\em demand transformer}
that transforms a {\em symbolic} demand  on its body to demands on its
arguments.  This summary  is then used to step  through function calls
during analysis.   Second, the concrete  demand on a function  body is
obtained through a 0-CFA-like conservative approximation that combines
the  demands on  all the  calls to  the function.   The result  of the
analysis  is the annotation  of each  program point  with finite-state
automata capturing  the liveness of variables at  the point. Depending
on the program point where  garbage collection has been triggered, the
GC consults  a set of automata  to curtail reachability
during marking.  This results in  an increase in the garbage reclaimed
and consequently in fewer garbage collections.

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order {\em  eager} language, a straightforward  extension of the
earlier  technique  is  not  possible  for  lazy  languages,  where  a
heap-allocated  object may either  be an  unevaluated expression  or a
closure. Since  data is  made live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined,  laziness complicates liveness
analysis itself.  In  addition, apart from evaluated data,  we want to
extend  liveness-based collection also  to closures---closures with no
future use should also  be collected.  This involves tracking closures
to  their points of  creation at  runtime from  where we  obtain their
liveness.  This  requires significant  modification to the  earlier GC
scheme.

{\color  {Myblue}Rewrite after the experiments:  Experiments  with a
single
  generation  copying  collector   confirm  the  expected  performance
  benefits.   Liveness-based  collection  results  in an  increase  in
  garbage reclaimed.   As a  consequence, there is  a decrease  in the
  number of  collections and a  (some quantification) decrease  in the
  memory size required  to run programs. Besides there  is a reduction
  in the overall execution time for some programs.}

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%% MOTIVATING EXAMPLE
%% \newcommand{\nilfigure}
%% {\scalebox{0.75}{
%% \psset{unit=1mm,nodesep=0mm,labelsep=0.5mm}
%% \begin{pspicture}(0,0)(1,1)
%% %\psgrid[xunit=1cm,yunit=1cm,gridwidth=.2pt,subgridwidth=.1pt,
%% subgriddiv=5,subgridcolor=gray,gridcolor=blue](0,0)(1,1)
%% \putnode{start}{origin}{0}{0}{}
%% \putnode{stop}{origin}{10}{10}{}
%% \ncline[offsetB=0,nodesepB=0,linewidth=.7]{-}{start}{stop} %here
%% \end{pspicture}
%% }}

\begin{figure}[t!]
  \psset{unit=1mm}
  \begin{pspicture}(90,48)(0,-48)
    %\psframe(90,48)(0,-48)
    \begin{tabular}{@{}c@{}}
      %\psframebox
          {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
            \UFL\
            \UNL{0} (\DEFINE\ (\length\  \pl)
	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
            (\PRIM\ 1\ (\CAR\  \pl))))
            \UNL{0}
	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
	    \listb
	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
            (\CDR\  \lista)
            \listb))))
            \UNL{0}
            \UNL{0} (\LET\ \px\
            $\leftarrow$\ (\CONS\ $5$
            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	    \UNL{2}
            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
            (\length\ \pz))))))
	  \end{uprogram}
      }}
      \\ 
      (a) Example program. \\  \\
%      \includegraphics[width=.45\textwidth]{motiv-example}
      \input{mem-graph} \\ 
      \renewcommand{\arraystretch}{.9}
      \begin{tabular}[t]{p{0.9\columnwidth}}
        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
        denotes a closure. Thick edges denote live links. Traversal
        stops at edges marked $\times$ during garbage collection.
      \end{tabular}
    \end{tabular} 
  \end{pspicture}
  \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

@@

The edges shown by thick arrows  are those which are live at $\pi$.  A
cell is marked  and preserved during garbage collection  only if it is
reachable from the  root set through a path of  live edges.  All other
cells can be  reclaimed.  Thus if a garbage  collection takes place at
$\pi$  with  the  heap  shown  in  Figure~\ref{fig:mot-example}(b),  a
liveness-based collector (LGC) will  collect only the cells referenced
by  $\pz$, $(\CDR~  \pz)$  and (the  cells  constituting) the  closure
referenced by $(\CDR~(\CDR~\pz))$ need to be marked and preserved.  In
contrast,  a reachability-based collector  (RGC) would  have preserved
all cells.

@@

{\color {Myblue}Rewrite at the end: 
Section~\ref{sec:defs} introduces the syntax and semantics of the
language used to illustrate our
analysis.
% Section~\ref{sec:operational} describes the operational semantics of
% the language.
Liveness analysis for this language is  described in
Section~\ref{sec:liveness}. This section also has by  a  sketch  of  a
correctness proof  relative  to  a  non-standard
semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
as   finite-state  automata.    Section~\ref{sec:experiments}  reports
experimental  results  and Section~\ref{sec:lgc-always-better}  proves
that a liveness based collector  can never do more garbage collections
than a reachability based collector.
% and Section~\ref{sec:conclusion} concludes the paper.
}
\section{The target language---syntax and semantics}
\label{sec:defs}
%%We let $x$,  $y$, $z$ range over variables,  $f$ over user-functions
%% and $p$ over primitive functions ($\CONS$, $\PRIM$ etc.).
Figure~\ref{fig:lang-syntax} describes the  syntax of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective}  where all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation  of closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x  \leftarrow s\,\,
\IN\,\, e$, $x$ may occur  in $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\  to a single
definition  is  for ease  of  exposition---generalization to  multiple
definitions  does  not add conceptual  difficulties.  We  further
restrict each variable  in a program to be distinct,  so that no scope
shadowing   occurs---this  simplifies   proofs   of  soundness.    The
$\SRETURN$ and \SIF\ expressions trigger evaluation of closures.

@@

% \paragraph{Inlining.}
{ \color {Myblue} Relook after proof: In spite of the  ANF restrictions
it
  is still possible  to inline non-recursive functions (a  fact we use
  to  prove the safety  of liveness  analysis).  A  user-function call
  $(\LET\,\, x \leftarrow (f\,y_1\,\ldots\,y_n)  \,\, \IN\,\, e)$ to a
  function  defined  (after renaming  its  formals  and  locals to  be
  disjoint         from         existing         variables)         by
  $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ e_f)$ is replaced by a sequence of
  $\LET$'s of  the form  $z_i\leftarrow (\ID\,\,y_i)$ followed  by the
  body $e_f$  but with  its $(\SRETURN\,\,w)$ expressions  replaced by
  $(\LET\,\, x \leftarrow (\ID\,\, w) \,\, \IN\,\, e)$.}

@@

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
Since all data objects are boxed, we model an environment as a mapping
from the set  of variables of the program  $\mathit{Var}$ to locations
in the  heap.  These  either contain  a value in  WHNF (a  number, the
empty  list  $\NIL$,  or  a  \CONS\  cell  with  possibly  unevaluated
constituents) or  a {\em  closure}.  A closure  is a pair  $\langle s,
\rho\rangle$  in which  $s$  is an  unevaluated  application, and  the
values of its free variables is given by $\rho$.

@@

A variable is {\em live} if  there is a possibility of its value being
used in  future computations  and dead if  it is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
liveness  of references.   Using  $\acar$, $\acdr$  to represent  access
using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
reachable from a  variable can be represented by a  set of {\em access
  paths} i.e.  prefix-closed  strings from $\{\acar,\acdr\}^\ast$.  As
an  example,  if $x$  is  a  list  with liveness  $\{\epsilon,  \acdr,
\acdr\acar,  \acdr\acdr, \acdr\acdr\acar\}$, then  future computations
can only  refer up  to the second  and third  members of $x$.   A {\em
  liveness  environment} is  a mapping  from variables  to  subsets of
$\{\acar,\acdr\}^\ast$, but  often expressed as a set,  for example by
writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$ instead of
$[x     \mapsto\{\epsilon,\acdr,\acdr\acdr\},    y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.    In  this   notation,  $x   \mapsto  \{\epsilon\}$
represents access using $x$ itself and $x \mapsto \{\}$ indicates that
$x$ is dead.  We  associate liveness environments with program points.
The liveness at $\pi:e$ is the liveness just before executing $e$.

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the future use of variables, a demand represents the future
use of the value of an  expression. The demand on an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\SRETURN~x)$.   Assume that  the  context of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon, y.\acdr, y.\acdr\acar \rbrace$ at $\pi$. These are the set
of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
\rbrace$  paths   of  the  result   of  $(\CDR~y)$\footnote{Reproduced
  from~\cite{asati14lgc}: In  classical strong liveness  analysis, $y$
  and $z$  are live at the  entry $n: x:=y+z$,  if and only if  $x$ is
  live  at exit  of $n$.   In our  terminology, the  liveness $\lbrace
  \epsilon\rbrace $ of $x$ at the  exit from $n$ becomes the demand on
  $y+z$, and this, in turn generates the liveness $\lbrace y.\epsilon,
  z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\
%% field of the cons cell pointed by
%% $y$.

Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves).
We use $\sigma$  to range over demands, $\alpha$  to range over access
paths  and $\Lv$ to  range over  liveness environments.   The notation
$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
abuse notation to  juxtapose an edge label and a  set of access paths:
$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$. A more
elaborate explanation of these notions appear in \cite{asati14lgc}

>>>>

\begin{abstract}
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs.  To do this, we  analyze programs for
liveness of  heap-allocated data.  The accompanying  garbage collector
is modified  to use the result  of the analysis to  preserve only live
data---a subset  of reachable data---during garbage  collection.  This
results  in an  increase  in the  garbage  reclaimed and  consequently
reduces the  memory footprint of  programs.  While this  technique has
already been shown to yield  benefits for eager first-order languages,
the lack of a statically determinable execution order and the presence
of closures at run-time pose additional challenges for lazy languages.
This requires changes  both in the earlier liveness  analysis and also
in the  garbage collection scheme  required to reclaim memory  that is
not live.

@@

\keywords{Heap Analysis, Liveness Analysis, Memory Management, Garbage
  Collection, Lazy Languages}

@@

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The analysis  has two  parts.  First,  each function  is
summarized in a context-sensitive manner as a {\em demand transformer}
that transforms a {\em symbolic} demand  on its body to demands on its
arguments.  This summary  is then used to step  through function calls
during analysis.   Second, the concrete  demand on a function  body is
obtained through a 0-CFA-like conservative approximation that combines
the  demands on  all the  calls to  the function.   The result  of the
analysis  is an  annotation of  each program  point with  finite-state
automata capturing the liveness of  variables at that point. Depending
on the program point where  garbage collection has been triggered, the
GC consults a set of  automata to restrict reachability during marking.
This results in an increase  in the garbage reclaimed and consequently
in fewer garbage collections.

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order {\em  eager} language, a straightforward  extension of the
earlier  technique  is  not  possible  for  lazy  languages,  where  a
heap-allocated   object   may   be  an   unevaluated   expression   or
closure. Since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.   In addition, apart  from evaluated data,  we extend
extend liveness-based  collection also to closures---closures  with no
future use should also be  collected.  This involves tracking closures
to their  points of  creation at  runtime from  where we  obtain their
liveness.  This  requires significant  modification to the  earlier GC
scheme.

{\color {Myblue}Rewrite  after the  experiments: Experiments
  with  a single  generation  copying collector  confirm the  expected
  performance  benefits.   Liveness-based  collection  results  in  an
  increase  in  garbage  reclaimed.   As a  consequence,  there  is  a
  decrease in  the number of  collections and a  (some quantification)
  decrease in the memory size  required to run programs. Besides there
  is a reduction in the overall  execution time for some programs. Our
  experiments further show that a mixed strategy, where we do liveness
  based GC  up to the closure  boundary and reachability based  GC for
  the  arguments (free  variables) of  the closure,  gives us  similar
  benefits as  a fully liveness based  GC (where the arguments  of the
  closure are also collected using liveness) at much less overhead (?).
}

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\begin{figure}[t!]
  \psset{unit=1mm}
  \begin{pspicture}(90,48)(0,-48)
    %\psframe(90,48)(0,-48)
    \begin{tabular}{@{}c@{}}
      %\psframebox
          {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
            \UFL\
            \UNL{0} (\DEFINE\ (\length\  \pl)
	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
            (\PRIM\ 1\ (\CAR\  \pl))))
            \UNL{0}
	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
	    \listb
	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
            (\CDR\  \lista)
            \listb))))
            \UNL{0}
            \UNL{0} (\LET\ \px\
            $\leftarrow$\ (\CONS\ $5$
            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	    \UNL{2}
            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
            (\length\ \pz))))))
	  \end{uprogram}
      }}
      \\ 
      (a) Example program. \\  \\
%      \includegraphics[width=.45\textwidth]{motiv-example}
      \input{mem-graph} \\ 
      \renewcommand{\arraystretch}{.9}
      \begin{tabular}[t]{p{0.9\columnwidth}}
        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
        denotes a closure. Thick edges denote live links. Traversal
        stops at edges marked $\times$ during garbage collection.
      \end{tabular}
    \end{tabular} 
  \end{pspicture}
  \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

@@

The edges shown by thick arrows are  those which are live at $\pi$.  A
cell is  preserved during garbage  collection only if it  is reachable
from the root set  through a path of live edges.   All other cells can
be reclaimed.  Thus if a garbage  collection takes place at $\pi$ with
the  heap shown  in Figure~\ref{fig:mot-example}(b),  a liveness-based
collector  (LGC) will  preserve only  the cells  referenced by  $\pz$,
$(\CDR~ \pz)$ and  (the cells constituting) the  closure referenced by
$(\CDR~(\CDR~\pz))$.   In  contrast,  a  reachability-based  collector
(RGC) would have preserved all cells.

@@

{\color {Myblue}Rewrite at the end: 
Section~\ref{sec:defs} introduces the syntax and semantics of the
language used to illustrate our
analysis.
% Section~\ref{sec:operational} describes the operational semantics of
% the language.
Liveness analysis for this language is  described in
Section~\ref{sec:liveness}. This section also has by  a  sketch  of  a
correctness proof  relative  to  a  non-standard
semantics.  Section~\ref{sec:computing}  shows how to  encode liveness
as   finite-state  automata.    Section~\ref{sec:experiments}  reports
experimental  results  and Section~\ref{sec:lgc-always-better}  proves
that a liveness based collector  can never do more garbage collections
than a reachability based collector.
% and Section~\ref{sec:conclusion} concludes the paper.
}
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  the  reasoning   about  the  program.   The
$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
evaluation of closures.

@@

@@

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

Here  $\mathit{Loc}$ is  a countable  set  of locations  in the  heap.
Since all data objects are boxed, we model an environment as a mapping
from the set  of variables of the program  $\mathit{Var}$ to locations
in the  heap.  These  either contain  a value in  WHNF (a  number, the
empty  list  $\NIL$,  or  a  \CONS\  cell  with  possibly  unevaluated
constituents) or  a {\em closure}.   A closure  is a pair  $\langle s,
\rho\rangle$ in  which $s$ is  an unevaluated application,  and $\rho$
maps free variables of $s$ to their respective locations.

@@

A variable is {\em live} if there  is a possibility of its value being
used in  future computations and  dead if  it is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
liveness of  references.  Using  $\acar$, $\acdr$ to  represent access
using  $\CAR$  and  $\CDR$  fields,  the  liveness  of  the  structure
reachable from a  variable can be represented by a  set of {\em access
  paths}     i.e.     prefix-closed     set     of    strings     from
$\{\acar,\acdr\}^\ast$.  As an example, if $x$ is a list with liveness
$\{\epsilon, \acdr,  \acdr\acar, \acdr\acdr,  \acdr\acdr\acar\}$, then
future computations can only refer up  to the second and third members
of $x$.  A  {\em liveness environment} is a mapping  from variables to
subsets of $\{\acar,\acdr\}^\ast$,  but often expressed as  a set, for
example by writing $\{x.\epsilon, x.\acdr, x.\acdr\acdr, y.\epsilon\}$
instead       of       $[x       \mapsto\{\epsilon,\acdr,\acdr\acdr\},
  y\mapsto\{\epsilon\}, z\mapsto\{\}]$.  In  this notation, $x \mapsto
\{\epsilon\}$ represents access using $x$  itself and $x \mapsto \{\}$
indicates that $x$  is dead.  We associate  liveness environments with
program points.  The  liveness at $\pi:e$ is the  liveness just before
executing $e$.

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the future use of variables, a demand represents the future
use of the value of an expression.  The demand on an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\SRETURN~x)$.  Assume  that  the  context of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes the  demand on
$(\CDR~y)$.    This,  in   turn,  generates   the  liveness   $\lbrace
y.\epsilon, y.\acdr, y.\acdr\acar \rbrace$ at $\pi$. These are the set
of $y$-rooted  accesses required  to explore $\lbrace  \epsilon, \acar
\rbrace$  paths  of  the  result of  $(\CDR~y)$.  In  strong  liveness
analysis       (dual      of       classical      faint       variable
analysis~\cite{horwitz.faint}),  $y$ and  $z$  are live  at the  entry
$\pi: x:=y+z$, if  and only if $x$  is live at exit of  $\pi$.  In our
terminology, the  liveness $\lbrace  \epsilon\rbrace $  of $x$  at the
exit from $n$ becomes the demand on $y+z$, and this, in turn generates
the liveness $\lbrace y.\epsilon, z.\epsilon  \rbrace$ at the entry of
$n$.

{
\color {Myblue}[I don't think this is valid anymore!!]
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves).
}

We follow the notations appearing in \cite{asati14lgc}. In particular,
We use $\sigma$  to range over demands, $\alpha$ to  range over access
paths and  $\Lv$ to  range over  liveness environments.   The notation
$\sigma_1\sigma_2$  denotes  the  set $\lbrace  \alpha_1\alpha_2  \mid
\alpha_1 \in \sigma_1, \alpha_2  \in \sigma_2\rbrace$.  Often we shall
abuse notation to  juxtapose an edge label and a  set of access paths:
$\acar\sigma$ is a shorthand for $\lbrace\acar\rbrace\sigma$.
